---
title: "Weather and Uber & Lyft Ridership"
author: "John Cruz"
date: "2023-04-25"
output:
  pdf_document: default
  html_document: default
  beamer_presentation: default
  powerpoint_presentation: default
  slidy_presentation: default
urlcolor: blue
---

## Introduction

The National Oceanic and Atmospheric Administration (['NOAA'](https://www.weather.gov/ama/heatindex#:~:text=The%20heat%20index%2C%20also%20known,sweat%20to%20cool%20itself%20off.)) defines the heat index as the apparent temperature of what the temperature feels like to the human body when relative humidity is combined with the air temperature. This has important considerations for the human body's comfort. When the body gets too hot, it begins to perspire or sweat to cool itself off.

As for the New York City subway system during the summer, it is notoriously known to have unbearable temperatures where the platform can be 104 degrees, compared to 86 degrees outside (['Curbed NY'](https://ny.curbed.com/2018/8/10/17674900/nyc-subway-summer-heat-temperatures)).

Given the health risks, and general discomfort during high heat index days, this project will look into alternative modes of transportation, particularly ridesharing companies such as Uber and Lyft.

---

## Research question

Does high heat index days (\>=90 degrees) increase the number of trips taken with Uber or Lyft compared to non-high heat index days?

---

## Data Source

**Weather ([Oikolab](https://oikolab.com/))**

Data was collected using [Oikolab API](https://docs.oikolab.com/#2-weather-data-weather) historical data API service. It collects its data from the ECWMF and NOAA. Each case represents hourly weather measurements in from 2021-2022.

**Uber & Lyft Trips ([NYC Taxi and Limousine Commission](https://www.nyc.gov/site/tlc/index.page))**

Data was collected using the available ['parquet files'](https://www.nyc.gov/site/tlc/about/tlc-trip-record-data.page). The agency collects the data from Uber and Lyft. Each case represents a trip taken either via Uber or Lyft between 2021-2022.

---

## Type of study

This is an observational study.

---

## Variables

**Dependent** 
- total trips: numerical

**Independent Variable(s)** 
The independent variables are:

-   type_of_day: categorical
-   precipitation: numerical
-   day_of_week: categorical
-   month: categorical
-   year: categorical

*Note:* Other potential factors that are important but not included: special events (i.e. sporting event), major delays with public transportation (MTA Subway) or alternative transportation such as Citi bikes.

---

## Required Libraries

```{r library, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(DBI)
library(lubridate)
library(weathermetrics)
library(infer)
library(psych)
```

---

## Data Preparation

### Load Historical Weather Data

**Calculate Heat Index**
-   Relative humidity is calculated using the temperature and dewpoint temperature.
-   Heat index is calculated using the temperature and relative humidity.

```{r read-data-weather, message=FALSE}
weather <- read_csv('data//oikolabs.csv') |> 
  janitor::clean_names()

weather <- 
  weather |> 
  mutate(temp_deg_f = celsius.to.fahrenheit(temperature_deg_c),
         rel_humidity = dewpoint.to.humidity(t = temperature_deg_c, 
                                             dp = dewpoint_temperature_deg_c, 
                                             temperature.metric = "celsius"),
         heat_idx = heat.index(t = temp_deg_f, 
                               rh = rel_humidity),
         total_precipitation_mm_of_water_equivalent = total_precipitation_mm_of_water_equivalent / 25.4) |> 
  rename(total_precip = total_precipitation_mm_of_water_equivalent)

weather <-
  weather |> 
  mutate(day_of_week = wday(datetime_utc, label = TRUE, week_start = 1, abbr = FALSE),
         day_of_week = as.factor(day_of_week),
         datetime_ny = with_tz(datetime_utc, "America/New_York")) |>
  relocate(datetime_ny) |> 
  select(datetime_ny, temp_deg_f, rel_humidity, heat_idx, total_precip) |> 
  filter(between(datetime_ny, as.Date("2021-01-01"),
         as.Date("2022-12-31")))

knitr::kable(head(weather), caption = 'Weather Data')
```

### Load Uber and Lyft Trips

The NYC Taxi and Limousine Commission provides a data dictionary ['here'](https://www.nyc.gov/assets/tlc/downloads/pdf/data_dictionary_trip_records_hvfhs.pdf). The rideshare app companies such as Uber is coded as (HV0003) and Lyft (HV0005).

-   Trips were filtered because of huge outliers that were present such as:
    -   Trip time had to be \>0 seconds and \<= 5 hours.
    -   Trip miles had to be \>= 0.
    -   Driver pay \> \$0.01.
    -   Base passenger fare \> \$0.01.
    -   Pickup locations had to be within the NYC region and not unknown/outside of it.
    
Given the large amounts of data to be processed, some of the data cleaning and filtering was done through DuckDB. DuckDB contains columnar-vectorized query execution engine where it allows for memory resources not to be severely depleted while trying to aggregate through the data. For more information visit ['DuckDB'](https://duckdb.org/why_duckdb.html#duckdbissimple)

```{r read-data-tlc, warning=FALSE, message=FALSE}
cnxn = dbConnect(duckdb::duckdb(), dbdir=":memory:")

dbExecute(cnxn, "CREATE VIEW tlc_trips AS 
                      SELECT 
                             CASE 
                                 WHEN hvfhs_license_num == 'HV0003' THEN 'Uber'
                                 WHEN hvfhs_license_num == 'HV0005' THEN 'Lyft'
                                 ELSE 'Other'
                             END AS app, 
                             pickup_datetime, 
                             dropoff_datetime, 
                             PULocationID, 
                             trip_miles, 
                             trip_time, 
                             base_passenger_fare
                      FROM 'data\\*.parquet'
                      WHERE 
                          trip_time >= 0 AND
                          trip_time < 18000 AND
                          trip_miles >= 0 AND
                          driver_pay > 0.01 AND
                          base_passenger_fare > 0.01 AND
                          PULocationID NOT IN (264, 265)"
          )

query <- "WITH floor_date AS(
                SELECT 
                  app,
                  time_bucket(interval '1 hour', pickup_datetime) AS pickup_datetime,
                  PULocationID, 
                  trip_miles, 
                  trip_time, 
                  base_passenger_fare
                FROM tlc_trips
          )
          
          SELECT 
            app,
            pickup_datetime,
            PULocationID,
            COUNT(*) as trips,
            SUM(trip_miles) AS trip_miles,
            SUM(trip_time) AS trip_time, 
            SUM(base_passenger_fare) AS base_passenger_fare
          FROM floor_date
          GROUP BY app, pickup_datetime, PULocationID
         "

db_trips <- dbGetQuery(cnxn, query)

tlc_trips <- 
  db_trips |> 
  mutate(pickup_datetime = force_tz(pickup_datetime, tzone = 'America/New_York'))

tlc_trips <- 
  tlc_trips |> 
  group_by(pickup_datetime) |> 
  summarise(total_trips = sum(trips),
            total_trip_dist = sum(trip_miles),
            total_trip_time = sum(trip_time),
            total_base_fare = sum(base_passenger_fare))
            
knitr::kable(head(tlc_trips), caption = 'Uber & Lyft Trips')
```

### Merge Datasets

The data from both sources will be merged together based on the datetime columns. It is important to note that due to the large size of source files, a parquet file of the cleaned data will be exported as *trip_weather_data.parquet* and provided in the repo. For the original files, access them from the previously mentioned methods. 

```{r merge-datasets}
trip_weather_data <- 
  tlc_trips |> 
  left_join(weather, by = join_by(pickup_datetime == datetime_ny)) |> 
  mutate(pickup_date = date(pickup_datetime)) |> 
  select(!pickup_datetime) |> 
  group_by(pickup_date) |> 
  mutate(total_trips = sum(total_trips),
         trip_dist = sum(total_trip_dist),
         trip_time = sum(total_trip_time),
         base_fare = sum(total_base_fare),
         temp_deg_f = max(temp_deg_f),
         heat_idx = max(heat_idx),
         precip = sum(total_precip),
         .keep = "none") |> 
  distinct() |> 
  mutate(type_of_day = case_when(heat_idx >= 90 ~ 'hot', 
                                 .default = 'normal'),
         day_of_week = factor(wday(pickup_date, label = TRUE, week_start = 1), ordered = FALSE),
         day_of_week = fct_relevel(day_of_week, "Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"),
         month = month(pickup_date, label = TRUE),
         year = year(pickup_date)) |> 
  drop_na()

trip_weather_data$day_of_week = relevel(trip_weather_data$day_of_week, ref='Mon')
trip_weather_data$type_of_day = relevel(factor(trip_weather_data$type_of_day, ordered = FALSE), ref='normal')
trip_weather_data$month = relevel(factor(trip_weather_data$month, ordered = FALSE), ref='Jan')
trip_weather_data$year = relevel(factor(trip_weather_data$year, ordered = FALSE), ref='2021')

write_parquet(trip_weather_data, "trip_weather_data.parquet")

knitr::kable(head(trip_weather_data))
```

--- 

## Summary Statistics

### Total Daily Trips
Here we can see the total number of daily trips taken via Uber and Lyft in NYC between 2021-2022. There is a seasonality trend occurring between the dates that we will look further into. 

```{r tlc-summary}
trip_weather_data |> 
  ggplot(aes(x = pickup_date, y = total_trips)) + 
  geom_line(stat = 'identity') +
  scale_y_continuous(labels = scales::comma) +
  theme_bw() +
  labs(title = "Total Daily Trips", caption = "Figure 1", x = '', y = '')
```

### Day of the Week vs. Total Trips
Plotting each year's total trips and the day of the week, we do see a seasonal trend where on average Monday's have the lowest trip counts and it progressively increases until Sunday's drop. 

```{r week}
ggplot(aes(x = day_of_week, y = total_trips), data = trip_weather_data) +
  geom_boxplot() +
  facet_grid(rows = vars(year(pickup_date))) +
  theme_bw() +
  labs(x = '', y = 'Total Trips', caption = 'Figure 2') +
  scale_y_continuous(labels = scales::comma)
```

### Daily Heat Index
Continuing looking into seasonal trends, we see both years have the same shape between the daily heat index and how it varies month to month. The high heat index months of interest is usually between late May until early August, which coincides with the summer months of NYC. 

```{r weather-summary}

trip_weather_data |> 
  ggplot(aes(x = month, y = heat_idx, color = type_of_day)) +
  geom_jitter(stat = 'identity') +
  facet_grid(rows = vars(year(pickup_date))) +
  theme_classic() +
  labs(x = '', y ='', title = "Daily Heat Index", caption = "Figure 3") +
  theme(legend.position = "top",
        legend.justification = c(0, 1)) +
  scale_color_manual(values = c('#4E79A7', '#F28E2B'), name = '')
```

### Heat Index vs. Daily Trips
Finally, let us look at the relationship of heat index to total number of trips. The scatter plot shows us that there is some bend initially as the heat index rises, but overall no strong trend. However, both years again follow the same seasonal patters. 

```{r plot-heat-trips}
trip_weather_data |> 
  ggplot(aes(x = heat_idx, y = total_trips, color = year)) + 
  geom_point(stat = 'identity') +
  labs(title = 'Heat Index vs. Total Trips', caption = "Figure 4", x = 'heat index', y = 'total trips') +
  theme_classic() +
  theme(legend.position = "top",
        legend.justification = c(0, 1)) +
  scale_color_manual(values = c('#5F9ED1', '#C85200'), name = '') +
  scale_y_continuous(labels = scales::comma)
  
```

### Observed Difference
Viewing the observed difference, we can see out of two years worth of data, there were 51 days where the heat index was over 90 degrees. However, on average the trips taken on hot days compared to normal days were slightly higher by about 6,000 trips. 

```{r observed-difference}
comparison <-
  trip_weather_data |> 
  group_by(type_of_day) |> 
  summarise(avg_trips = mean(total_trips),
            n = n())

knitr::kable(comparison, caption = "Observed Differences")
```

--- 

## Hypothesis Testing
To test the observed differences between both high heat index days and non-high heat index days, a hypothesis test will be performed. To account for the possible differences in the two year data, a bootstrap was performed to account for such variability. 

```{r set-seed}
set.seed(2000)
```

```{r hypothesis-setup}
obs_diff <- trip_weather_data %>%
  specify(total_trips ~ type_of_day) %>%
  calculate(stat = "diff in means", order = c("hot", "normal"))
```

```{r null-dist}
null_dist <- trip_weather_data %>%
  specify(total_trips ~ type_of_day) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("hot", "normal"))
```

We can see through the histogram distribution, that our data is normally distributed.

```{r plot-hist}
ggplot(data = null_dist, aes(x = stat)) +
  geom_histogram(bins = 30) +
  labs(caption = "Figure 5")
```

After confirming the normality of the distribution a two-tailed t-test will be performed.

```{r t-test}
t.test(total_trips ~ type_of_day, data = trip_weather_data)
```

Using a 95% confidence t-test, it tells us that we are unable to reject the null hypothesis and claim that a high heat index increases the total daily trips taken using Uber or Lyft. We see in the confidence interval, it includes zero, where as the p-value is also 0.6675 which is />0.05 alpha threshold. 

--- 

## Alternative Heat Index Model
While we failed to reject the null hypothesis, using a t-test, there are other controlling variables to account for that may influence how a heat index may be prove to be useful. A multiple linear regression model will be created to account for the previous seasonal variability accounted for in our previous figures. *day_of_week* will factor out each day of the week, having Monday as the reference variable, *month* will factor out each month of the year, having January as the reference variable and *year*, factoring out for the two year data collected, having 2021 as the reference variable. Lastly, daily total precipitation will be added to account for potential variation. 

### Precipitation
Here we can see no linear trend of precipitation to total daily trips. 

```{r daily-precip}
ggplot(aes(x = precip, y = total_trips), data = trip_weather_data) +
  geom_point() + 
  labs(x = 'precipitation (inches)', y = 'total trips') +
  scale_y_continuous(labels = scales::comma)
```

As for the correlation between heat index and precipitation we get 0.01965 which shows almost zero correlation with each other. 

```{r corr}
cor(trip_weather_data$heat_idx, trip_weather_data$precip)
```


### Model
Now we can build the linear regression as followed:

$$ 
\hat{y} = \beta_0 + x\beta_1 + x\beta_2 + x\beta_3 + x\beta_4 + x\beta_5
$$

where 

$$
\beta_1 = type\_of\_day\\
\beta_2 = precip\\
\beta_3 = day\_of\_week\\
\beta_4 = month\\
\beta_5 = year\\
$$

```{r model}
lm_mod <- lm(total_trips ~ type_of_day + precip + day_of_week + month + year, data = trip_weather_data)
summary(lm_mod)
```

### Results

The linear regression model

$$
\begin{aligned}
\hat{y} = 296687 
&\ + 16807 \times type\_of\_dayhot \\
&\ - 3264  \times daily\_precip \\
&\ + 18886 \times day\_of\_weekTue \\
&\ + 47471 \times day\_of\_weekWed \\
&\ + 77662 \times day\_of\_weekThu \\
&\ + 135131 \times day\_of\_weekFri \\
&\ + 172810 \times day\_of\_weekSat \\
&\ + 80637 \times day\_of\_weekSun \\
&\ + 65816 \times monthFeb \\
&\ + 103416  \times monthMar \\
&\ + 99763 \times monthApr \\
&\ + 104052 \times monthMay \\
&\ + 117217 \times monthJun \\
&\ + 85819 \times monthJul \\
&\ + 82951 \times monthAug \\
&\ + 116278 \times monthSep \\
&\ + 147176 \times monthOct \\
&\ + 141280 \times monthNov \\
&\ + 141938 \times monthDec \\
&\ + 103358 \times year2022 \\
\end{aligned}
$$

When controlling for the seasonality, the heat index (*type_of_dayhot*) became statistically significant. Holding all other variables constant, when the heat index is />=90 we can estimate an increase of 16807 trips than non high heat index days. As for the practical significance of the model, it is quite surprisingly a significant predictor as the adjusted $R^2$ of the model is 0.8244. 

---

## Checking Assumptions

**Normality** 
The Q-Q plot shows there is some "S" curvature within the band of residuals, but overall is straight. 

```{r qq-plot}
ggplot(data = lm_mod, aes(sample = .resid)) +
  stat_qq() +
  labs(caption = "Figure 5")
```

**Constant variability** 
The spread around zero does appear to have some heteroskedasticity as it is cone-shaped from the middle but overall nothing alarming. 

```{r residuals}
ggplot(data = lm_mod, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals") +
  labs(caption = "Figure 6")
```

**Linearity** 
It does not pass the linearity test as from Figure 3, there is not a linear shape for heat index and trips taken. 

---

## Conclusions

When using the heat index as a measure of trips taken, it alone does not seem to be an indicator of higher trips being taken via Uber and Lyft. However, when controlling for the seasonal effects on the day of the week, it does become a good indicator of trip counts. However, there are some concerns about the high results of the adjusted R^2 and small scale of data used. For future research and without limitations to processing capacity, I would want to increase the data from being just using August 2022 and study the trends for a few years worth of summer data. It is also important to note that this analysis focuses on the unique environment of NYC but can the same be said for other hot climates such as Miami or Los Angeles. 