---
title: "tlc_data_filter"
author: "John Cruz"
date: "2023-05-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Load Uber and Lyft Trips**

The NYC Taxi and Limousine Commission provides a data dictionary ['here'](https://www.nyc.gov/assets/tlc/downloads/pdf/data_dictionary_trip_records_hvfhs.pdf). The rideshare app companies such as Uber is coded as (HV0003) and Lyft (HV0005).

-   All the trips performed in August were found to be within two separate files for both August and September.
-   Trips were filtered because of huge outliers that were present such as:
    -   Trip time had to be \>0 seconds and \<= 5 hours.
    -   Trip miles had to be \>= 0.
    -   Driver pay \> \$0.01.
    -   Base passenger fare \> \$0.01.
    -   Pickup locations had to be within the NYC region and not unknown/outside of it.

```{r library, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(lubridate)
library(DBI)
```

```{r read-data-tlc, warning=FALSE, message=FALSE}
cnxn = dbConnect(duckdb::duckdb(), dbdir=":memory:")

dbExecute(cnxn, "CREATE VIEW tlc_trips AS 
                      SELECT 
                             CASE 
                                 WHEN hvfhs_license_num == 'HV0003' THEN 'Uber'
                                 WHEN hvfhs_license_num == 'HV0005' THEN 'Lyft'
                                 ELSE 'Other'
                             END AS app, 
                             pickup_datetime, 
                             dropoff_datetime, 
                             PULocationID, 
                             trip_miles, 
                             trip_time, 
                             base_passenger_fare
                      FROM 'data\\*.parquet'
                      WHERE 
                          trip_time >= 0 AND
                          trip_time < 18000 AND
                          trip_miles >= 0 AND
                          driver_pay > 0.01 AND
                          base_passenger_fare > 0.01 AND
                          PULocationID NOT IN (264, 265)"
          )

query <- "WITH floor_date AS(
                SELECT 
                  app,
                  time_bucket(interval '1 hour', pickup_datetime) AS pickup_datetime,
                  PULocationID, 
                  trip_miles, 
                  trip_time, 
                  base_passenger_fare
                FROM tlc_trips
          )
          
          SELECT 
            app,
            pickup_datetime,
            PULocationID,
            SUM(trip_miles) AS trip_miles,
            SUM(trip_time) AS trip_time, 
            SUM(base_passenger_fare) AS base_passenger_fare
          FROM floor_date
          GROUP BY app, pickup_datetime, PULocationID
         "

db_trips <- dbGetQuery(cnxn, query)

```

```{r mutate-data, message=FALSE}
tlc_trips <- 
  db_trips |> 
  mutate(pickup_datetime = force_tz(pickup_datetime, tzone = 'America/New_York'))

# removed app from group by
tlc_trips <- 
  tlc_trips |> 
  group_by(pickup_datetime) |> 
  summarise(total_trips = n(),
            total_trip_dist = sum(trip_miles),
            total_trip_time = sum(trip_time),
            total_base_fare = sum(base_passenger_fare))

# write_parquet(tlc_trips, 'tlc_trips.parquet')
```

---

## Previous Code

tlc_trips_aug <- read_parquet('fhvhv_tripdata_2022-08.parquet')

tlc_trips_aug <- 
  tlc_trips_aug |>
  mutate(app = ifelse(hvfhs_license_num == 'HV0003', 'Uber', 'Lyft')) |> 
  relocate(app) |> 
  filter(pickup_datetime > '2022-07-31 23:59:59',
         trip_time >= 0,
         trip_time < 18000,
         trip_miles >= 0,
         driver_pay > 0.01,
         base_passenger_fare > 0.01,
         !PULocationID %in% c(264, 265)) |> 
  select(app, pickup_datetime, dropoff_datetime, PULocationID, trip_miles, trip_time, base_passenger_fare)

tlc_trips_sep <- read_parquet('fhvhv_tripdata_2022-09.parquet')

tlc_trips_sep <- 
  tlc_trips_sep |>
  mutate(app = ifelse(hvfhs_license_num == 'HV0003', 'Uber', 'Lyft')) |> 
  relocate(app) |>
  filter(pickup_datetime < '2022-09-01 00:00:00',
         trip_time >= 0,
         trip_time < 18000,
         trip_miles >= 0,
         driver_pay > 0.01,
         base_passenger_fare > 0.01,
         !PULocationID %in% c(264, 265)) |> 
  select(app, pickup_datetime, dropoff_datetime, PULocationID, trip_miles, trip_time, base_passenger_fare)

tlc_trips <- 
  rbind(tlc_trips_aug, tlc_trips_sep)

tlc_trips <- 
  tlc_trips |> 
  mutate(pickup_datetime = force_tz(pickup_datetime, tzone = 'America/New_York'),
         pickup_floor = floor_date(pickup_datetime, unit = "hour"))

tlc_trips <- 
  tlc_trips |> 
  group_by(app, pickup_floor) |> 
  summarise(total_trips = n(),
            avg_trip_dist = mean(trip_miles),
            avg_trip_time = mean(trip_time),
            avg_base_fare = mean(base_passenger_fare))

write_parquet(tlc_trips, 'tlc_trips.parquet')